---
title: "Finches_EDA"
output: html_document
date: "2023-10-23"
---

```{r Read table, include=FALSE}
# Load the required package
library(readxl)
finches_file <- "C:/Users/Aitana/Downloads/Finches_Dataset_Evolution.xls"
finches_table <- read_excel(finches_file)
```


```{r Missing map plot, include=FALSE}

library(Amelia)
png("missmap.png", width = 800, height = 800)
missmap <- missmap(finches_table)
dev.off()


```


```{r cars}
library(dplyr)

# Replace spaces with periods in column names
colnames(finches_table) <- gsub(" ", ".", colnames(finches_table))
colnames(finches_table) <- gsub("\\(", ".", colnames(finches_table))
colnames(finches_table) <- gsub("\\)", ".", colnames(finches_table))



numeric_cols <- sapply(finches_table, is.numeric)
finches_table_numeric <- select_if(finches_table, is.numeric)
```

# 1. Description of the dataset: 


```{r TESTING NORMALITY OF THE VARIABLES}


shapiro_results <- sapply(finches_table_numeric, function(col) {
  if (is.numeric(col)) {
    shat <- shapiro.test(col)
    c(W = shat$statistic, p_value = shat$p.value)
    
  } else {
    NA  # Non-numeric columns will have NA results
  }
})

# Print the results
print(shapiro_results)
significance_level <- 0.05
normally_distributed <- names(which(sapply(shapiro_results, function(x) x["p_value"]) > significance_level)) #If the p-value is less than 0.05 then 
#the null hypotheses that the data is normally distributed is rejected. If p-value is > 0.05 then the null hypothesis is not rejected (its normally distributed)
cat("Normally distributed variables: ", paste(normally_distributed, collapse = ", "))

```


```{r CREATE DESCRIPTIVE SUMMARIES FOR ALL VARIABLES, echo=FALSE}
library(purrr)
library(moments)
library(gtsummary)
library(ggplot2)
library(jpeg)
library(xtable)
library(tidyverse)
library(tidyr)
library(forcats)


#Firstly, I created a summary of numeric variables including measures such as median, kurtosis, skewness...
#Then I used library gtsummary, an R library that makes descriptive summary tables from the data, in which 
#we have to indicate the measures we want to use to represent each variable. I saved the output to latex tables


#Descriptive summary of the data. 
df1 <- finches_table_numeric %>% map_dbl(mean)
df2 <- finches_table_numeric %>% map_dbl(median)
df3 <- finches_table_numeric %>% map_dbl(sd)
df4 <- finches_table_numeric %>% map_dbl(kurtosis)
df5 <- finches_table_numeric %>% map_dbl(skewness)
df6 <- finches_table_numeric %>% map_dbl(min)
df7 <- finches_table_numeric %>% map_dbl(max)

params <- rbind(df1, df2, df3, df4, df5, df6, df7)
rownames(params) <- c("Mean", "Median", "Sd", "Kurtosis", "Skewness", "Min", "Max")
params <- t(params)
params


new_tbl_numeric <- finches_table %>%
select(Weight..g., Wing..mm., Tarsus..mm., Beak.Length..mm., Beak.Depth..mm., Beak.Width..mm.)


new_tbl_categorical <- finches_table %>%
select(Sex, First.adult.year, Last.Year,Sex, Species)

tbl_num <- new_tbl_numeric %>%
  tbl_summary(
   
    statistic = list(
      "Weight..g." ~ "{median} ({IQR})",
      "Wing..mm." ~ "{median} ({IQR})", 
      "Tarsus..mm." ~ "{mean} ({sd})", 
      "Beak.Width..mm." ~ "{mean} ({sd})",
      "Beak.Depth..mm." ~ "{mean} ({sd})",
      "Beak.Length..mm." ~ "{mean} ({sd})"
    )
  )%>%
  as_gt() %>%
  gt::as_latex() %>%
  writeLines(con = "descriptive_summary_numeric.tex")



tbl_cat <- tbl_summary(new_tbl_categorical) %>%
  as_gt() %>%
  gt::as_latex() %>%
  writeLines(con = "descriptive_summary_cat.tex")

```

```{r Save categorical variable descriptive plots in png files, echo=FALSE}

library(gridExtra)
library(ggpubr)
library(vcd)



# Create individual plots for each variable with a single fill color
fill_color <- "#69b3a2"  # Specify the fill color here

plot1 <- ggplot(new_tbl_categorical, aes(x = Sex)) +
  geom_bar(fill = fill_color) +
  labs(title = "Sex")

plot2 <- ggplot(new_tbl_categorical, aes(x = First.adult.year)) +
  geom_bar(fill = fill_color) +
  labs(title = "First Year")

plot3 <- ggplot(new_tbl_categorical, aes(x = Last.Year)) +
  geom_bar(fill = fill_color) +
  labs(title = "Last Year")

new_tbl_categorical$survivor <- ifelse(new_tbl_categorical$Last.Year > 1977, "Survived", "Did not survive")
plot4 <- ggplot(new_tbl_categorical, aes(x = survivor)) +
  geom_bar(fill = fill_color) +
  labs(title = "Survivor")

# Arrange the plots in a 2x2 grid
grid_plot <- grid.arrange(plot1, plot2, plot3, plot4, ncol = 2)

# Save the entire grid as an image (e.g., "plots.png")
ggsave("plots.png", grid_plot, width = 8, height = 6, units = "in")

```

```{r Save numeric descriptive plots to png, fig.width=6, fig.height=5, out.width='600px' }

p <- new_tbl_numeric %>% pivot_longer(everything(), names_to = "cols", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins=20,fill="#69b3a2", color="#e9ecef") +
  facet_wrap(~ cols, ncol=3, scales = 'free', ) + 
  labs(title="Continuous variable distribution", x="Value", y="Nº of observations")


ggsave("numeric_distributions.png", p)
```




```{r Create boxplots for numeric variables }


p <- new_tbl_numeric %>%
  pivot_longer(everything(),names_to="cols",values_to="value")%>%
  ggplot(aes(y = value)) +
  geom_boxplot(outlier.colour="red",
  outlier.shape=19,
  outlier.size=1.5, 
  fill="#86ebd0",
  outlier.alpha = 0.3)+
  facet_wrap(~ cols, ncol=4, scales = 'free')

ggsave("boxplots.png", p)
```

## 2. Correlation analysis: pair plots and correlation matrix to asess relationships between morphological variables: 


## 2.1. Scatter plots
```{r correl scatter, fig.width=7, fig.height=7, out.width='6500px'}
pdf("pairs_plot.pdf", width = 8, height = 8)
png("pairs.png", width = 800, height = 800)

pairs(new_tbl_numeric[, !colnames(new_tbl_numeric) %in% "survivor"], pch = 16, col = "#69b3a2", cex = 0.3)
dev.off()
```


## 2.2. Correlation matrix
```{r correlation matrix with spearman coefficients}
library(rstatix)
cor.mat <- new_tbl_numeric %>% cor_mat(-survivor)
cor.mat %>% cor_get_pval()

png("corr_plot_pval.png", width = 800, height = 800)

cor.mat %>%
  cor_reorder() %>%
  pull_lower_triangle() %>%
  cor_plot(label = TRUE)
cor.mat %>% cor_gather()

```


- Weight (g) and Wing (mm): 0.648
- Weight (g) and Beak Length (mm): 0.851
- Weight (g) and Beak Depth (mm): 0.731
- Weight (g) and Beak Width (mm): 0.768
- Wing (mm) and Beak Length (mm): 0.780
- Wing (mm) and Beak Depth (mm): 0.648
- Wing (mm) and Beak Width (mm): 0.605
- Beak Length (mm) and Beak Depth (mm): 0.851
- Beak Length (mm) and Beak Width (mm): 0.744
- Beak Depth (mm) and Beak Width (mm): 0.878


### 3. Hypothesis testing: 

#### 3.1. HYPOTHESIS 1: For the first hypothesis, we seek to study the relationship between survival status and different morphological characteristics related to the beak: In these graphs, we also see the relationship between other morphological variables and survival (hypothesis 2)


```{r exper8, fig.width=3, fig.height=3, out.width='300px'}
library(ggplot2)
library(ggthemes)
#library(ggpubr) 
#library(cowplot)

finches_table_numeric$survivor <- ifelse(finches_table_numeric$Last.Year > 1977, "Survived", "Did not survive")
finches_table_numeric$Beak.Size <- with(finches_table_numeric, Beak.Depth..mm. * Beak.Length..mm. * Beak.Width..mm.)

# Create density plots for continuous variables against 'Survivor'
continuous_vars <- colnames(finches_table_numeric)[c(4:9, 11)]
# Set a custom color palette
my_colors <- c("Survived" = "skyblue", "Did not survive" = "orange")

# Create a PDF device to save the plots
```


```{r SAVE BIVARIATE PLOTS TO PDF, fig.width=3, fig.height=3, out.width='300px'}
pdf("density_box_plots.pdf", width = 10, height = 8)

for (var in continuous_vars) {
  # Create density plot
  density_plot <- ggplot(finches_table_numeric, aes(x = .data[[var]], fill = survivor)) +
    geom_density(alpha = 0.6) +
    labs(title = paste("Density Plot of", var, "vs. Survivor")) +
    scale_fill_manual(values = my_colors) +
    theme_minimal() +
    theme(legend.position = "top")
  
  # Create box plot
  box_plot <- ggplot(finches_table_numeric, aes(x = survivor, y = .data[[var]], fill = survivor)) +
    geom_boxplot() +
    labs(title = paste("Box Plot of", var, "vs. Survivor")) +
    scale_fill_manual(values = my_colors) +
    theme_minimal() +
    theme(legend.position = "top")
  
  # Arrange density and box plots side by side
  plot_arranged <- plot_grid(density_plot, box_plot, ncol = 2)
  
  # Print the arranged plot
  print(plot_arranged)
}

# Close the PDF device
dev.off()


```









```{r SAVE BIVARIATE PLOTS TO PNG FILES}

# Create a directory to save the PNG files
dir.create("density_box_plots")

# Set the custom color palette
my_colors <- c("Survived" = "skyblue", "Did not survive" = "orange")

for (var in continuous_vars) {
  # Create density plot
  density_plot <- ggplot(finches_table_numeric, aes(x = .data[[var]], fill = survivor)) +
    geom_density(alpha = 0.6) +
    labs(title = paste("Density Plot of", var, "vs. Survivor")) +
    scale_fill_manual(values = my_colors) +
    theme_minimal() +
    theme(legend.position = "top")
  
  # Create box plot
  box_plot <- ggplot(finches_table_numeric, aes(x = survivor, y = .data[[var]], fill = survivor)) +
    geom_boxplot() +
    labs(title = paste("Box Plot of", var, "vs. Survivor")) +
    scale_fill_manual(values = my_colors) +
    theme_minimal() +
    theme(legend.position = "top")
  
  # Arrange density and box plots side by side
  plot_arranged <- plot_grid(density_plot, box_plot, ncol = 2)
  
  # Save the plot as a PNG file
  ggsave(filename = paste("density_box_plots/", var, ".png", sep = ""), plot = plot_arranged, width = 10, height = 8)
}



```
#### 3.1.2. P-values and summary table


```{r COMPUTE THE P-VALUES FOR THE OBSERVED RELATIONSHIPS IN ORDER TO TEST THE FIRST HYPOTHESIS}
finches_table_numeric %>%
  select("Beak.Width..mm.", "Beak.Depth..mm.", "Beak.Length..mm.", "survivor") %>%
  tbl_summary(
    by = survivor,
    statistic = list(
      "Beak.Width..mm." ~ "{mean} ({sd})",
      "Beak.Depth..mm." ~ "{mean} ({sd})",
      "Beak.Length..mm." ~ "{mean} ({sd})"
    )
  ) %>%
  add_p(
    test = list(
      "Beak.Width..mm." ~ "t.test",
      "Beak.Depth..mm." ~ "t.test",
      "Beak.Length..mm." ~ "t.test"
    
    ),
    pvalue_fun = function(x) style_number(x, digits = 3)
  ) %>% as_gt() %>%
  gt::as_latex() %>% writeLines(con = "latex_out_pvals_beaks_natural_selec.tex")

```

### 2. HYPOTHESIS 2 (p.values): Having already observed the relationship between the morphological variables, we now want to check if there's significant relationship between other morphological variables and survival: 


```{r SAVE BIVARIATE SUMMARY TABLE TO LATEX (morphological)}
finches_table_numeric %>%
  select("Weight..g.", "Wing..mm.", "Tarsus..mm.", "survivor") %>%
  tbl_summary(
    by = survivor,
    statistic = list(
      "Weight..g." ~ "{median} ({IQR})", 
      "Wing..mm." ~ "{median} ({IQR})", 
      "Tarsus..mm." ~ "{mean} ({sd})"
    )
  ) %>%
  add_p(
    test = list(
      "Weight..g." ~ "wilcox.test", 
      "Wing..mm." ~ "wilcox.test", 
      "Tarsus..mm." ~ "t.test"
    ),
    pvalue_fun = function(x) style_number(x, digits = 3)
  ) %>% as_gt() %>%
  gt::as_latex() %>% writeLines(con = "pvals_morf_seleccion_natural.tex")


```



## 3.3. HYPOTHESIS 3: There is sexual dimorphism after the drought: 

#### 3.3.1. We check normality again in the reduced sample (only survivors of the drought)
```{r Saphiro for survivors to the drougth to test normality in the new sample (reduced sample)}

# Assuming finches_table_numeric is a data frame
shapiro_results <- sapply(finches_table_numeric[finches_table_numeric$survivor == "Survived",], function(col) {
  if (is.numeric(col)) {
    shat <- shapiro.test(col)
    c(W = shat$statistic, p_value = shat$p.value)
    
  } else {
    NA  # Non-numeric columns will have NA results
  }
})

# Print the results
print(shapiro_results)
significance_level <- 0.05
normally_distributed <- names(which(sapply(shapiro_results, function(x) x["p_value"]) > significance_level)) #If the p-value is less than 0.05 then 
#the null hypotheses that the data is normally distributed is rejected. If p-value is > 0.05 then the null hypothesis is not rejected (its normally distributed)
cat("Normally distributed variables: ", paste(normally_distributed, collapse = ", "))


```



#### 3.3.2. (Not included in the study): is survival affected by sex?? Bivariate analysis and descriptive tables
```{r IS SURVIVAL AFFECTED AT ALL BY SEX?}

finches_table_numeric$Sex <- finches_table$Sex
filtered_data <- finches_table_numeric[finches_table_numeric$Sex %in% c("male", "female"), ]


ggsave("bar_plot_surv_sex.png", bar_plot, width = 6, height = 4)
bar_plot <- ggplot(filtered_data, aes(x = Sex, fill = survivor)) +
  geom_bar(position = "dodge") +
  labs(title = "Survivor by Sex") +
  scale_fill_manual(values = c("Survived" = "skyblue", "Did not survive" = "orange")) +
  theme_minimal() +
  theme(legend.position = "top")

# Print the bar plot
print(bar_plot)

# Create a table summary
table_summary <-  filtered_data %>% select("Sex", "survivor") %>% tbl_summary(by = Sex) %>%
  add_p() %>%
    as_gt() %>%
  gt::as_latex() %>%
  writeLines(con = "latex_out_sexo_drought.tex")

print(table_summary)
```



#### 3.3.3. Visualizing the change in beak size for males and females (before and after the drought)

```{r OBSERVE THE CHANGE BEFORE - AFTER THE DROUGHT IN FEMALES AND MALES }

males <- finches_table_numeric[finches_table_numeric$Sex == "male", ]
females <- finches_table_numeric[finches_table_numeric$Sex == "female", ]

# Set the size of the boxplots
par(mfrow = c(2, 1), mar = c(5, 5, 4, 2))

# Create a directory to save PNG files
dir.create("boxplots", showWarnings = FALSE)

# Boxplot for beak depth in males comparing survivors and non-survivors
png("boxplots/male_beak_depth.png", width = 800, height = 600)
boxplot(Beak.Depth..mm. ~ survivor, data = males, 
      
        main = "Beak Depth in Males",
        col = c("orange", "lightblue"))
dev.off()

# Boxplot for beak depth in females comparing survivors and non-survivors
png("boxplots/female_beak_depth.png", width = 800, height = 600)
boxplot(Beak.Depth..mm. ~ survivor, data = females, 
  
        main = "Beak Depth in Females",
        col = c("orange", "lightblue"))
dev.off()



```

#### 3.3.4. Visualizing differences in morphological characteristics in survivors of the drought by sex to test hypothesis 3

```{r PLOT DIFFERENCES BETWEEN MALES AND FEMALES AFTER THE DROUGHT}
# Define the output directory for PNG files
output_dir <- "sex_bivariate"

# Create the output directory if it doesn't exist
dir.create(output_dir, showWarnings = FALSE)

# Set the working directory to the output directory
setwd(output_dir)

finches_table_numeric_surv <- finches_table_numeric[finches_table_numeric$survivor == "Survived" & finches_table_numeric$Sex != "unknown", ]

# Define the file extension (e.g., "png")
file_extension <- "png"

# Define custom color palette
my_colors <- c("male" = "steelblue", "female" = "salmon")

# Loop through the variables and save plots as PNG files
for (var in continuous_vars) {
  # Create density plot
  density_plot <- ggplot(finches_table_numeric_surv, aes(x = .data[[var]], fill = Sex)) +
    geom_density(alpha = 0.6) +
    labs(title = paste("Density Plot of", var, "vs. Sex")) +
    scale_fill_manual(values = my_colors) +
    theme_minimal() +
    theme(legend.position = "top")
  
  # Create box plot
  box_plot <- ggplot(finches_table_numeric_surv, aes(x = Sex, y = .data[[var]], fill = Sex)) +
    geom_boxplot() +
    labs(title = paste("Box Plot of", var, "vs. Sex")) +
    scale_fill_manual(values = my_colors) +
    theme_minimal() +
    theme(legend.position = "top")
  
  # Arrange density and box plots side by side
  plot_arranged <- plot_grid(density_plot, box_plot, ncol = 2)
  
  # Define the file name and path
  file_name <- paste(var, "vs_Sex_surv", file_extension, sep = ".")
  
  # Save the plot as a PNG file
  ggsave(filename = file_name, plot = plot_arranged, width = 10, height = 8, device = file_extension)
}

# Reset the working directory to its original location
setwd("C:/Users/Aitana/OneDrive")
```


#### 3.3.5. Finally, we compute p-values in order to observe if there are any significant relationships. 

```{r}

finches_table_numeric[finches_table_numeric$Sex %in% c("male", "female") & finches_table_numeric$survivor == "Survived", ] %>%
  select("Weight..g.", "Wing..mm.", "Tarsus..mm.", "Beak.Width..mm.", "Beak.Depth..mm.", "Beak.Length..mm.", "Sex") %>%
  tbl_summary(
    by = Sex,
    statistic = list(
      "Weight..g." ~ "{median} ({IQR})", 
      "Wing..mm." ~ "{mean} ({sd})", 
      "Tarsus..mm." ~ "{mean} ({sd})", 
      "Beak.Width..mm." ~ "{mean} ({sd})",
      "Beak.Depth..mm." ~ "{mean} ({sd})",
      "Beak.Length..mm." ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    )
  ) %>%
  add_p(
    test = list(
      "Weight..g." ~ "wilcox.test", 
      "Wing..mm." ~ "t.test", 
      "Tarsus..mm." ~ "t.test", 
      "Beak.Width..mm." ~ "t.test",
      "Beak.Depth..mm." ~ "t.test",
      "Beak.Length..mm." ~ "t.test"
    ),
    pvalue_fun = function(x) style_number(x, digits = 3)
  ) %>%
  as_gt() %>%
  gt::as_latex() %>%
  writeLines(con = "latex_out_pvals_dimorfismo_sexual_survived.tex")


```


#Extra hypothesis: not included in the report--> age of the finches at the moment of the drought affected their survival?
```{r exper8, fig.width=3, fig.height=3, out.width='300px}


summary_table <- tbl_summary(data,
  by = survivor,
  statistic = list(
    age_at_drought ~ "{mean} ({sd})"
  ),
  type = list(age_at_drought ~ "continuous")
)

# Add p-values based on the Wilcoxon test
summary_table <- summary_table %>%
  add_p()

# Print the summary table
summary_table

```
